### 2013-01-24 patchy-selection-paper.tex
/* Patchy Selection paper */

\section{General model}

outline here assumptions common to both models.

Stick to constant-speed waves.
Note monty's empircal speeds paper.

\subsection{Relation to a continuous model}
\label{ss:patchyspace}

A similar situation of interest is the case when the species occupies a continuous range, but the selective pressure itself is patchy.
If the areas where the new allele is advantageous are relatively isolated,
we might hope to apply the results above to such a situation.
For this purpose, suppose that instead of a uniform habitat,
the selective pressure is such that the new alleles are only advantageous in isolated locations, 
and are disadvantageous in the intervening region.
As before, we assume that the new allele has fitness $1+s_b$ relative to the unmutated type in the ``good'' patches,
and assume that it has fitness $1+s_m$ in the intervening areas, with $s_m<0$.
To make things tractable, we assume that the distances between the patches are large relative to migration distance $\sigma$,
and that the disadvantage $s_m$ ensures that on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
With these assumptions as justification, we will move to a model of discrete demes exchanging migrants,
where ``migration'' occurs through the rare, transient passage of the new allele through regions where it is disadvantageous.

equal deme sizes?  general migration rates?

\subsubsection{Mutational influx}
\label{ss:patchymutation}

Consider first a single, isolated patch of area $A$ in which no new allele has appeared.
\citet{barton1987establishment} (or \citet{slatkin1973geneflow}?) has studied this situation (in one dimension)
and found that:
mutations appearing further than a few multiples of $\sigma/\sqrt{2|s_m|}$ away from the patch have vanishingly small probability of establishment;
mutations appearing within the patch have probability strictly less but approaching that $2s_b$ of establishment;
and if the patch width is less than $2 \tan^{-1} (\sqrt{|s_m/s_b|})$, then local fixation does not occur (rather, is not stable).
Let $p(x)$ be the probability that a new allele arising at location $x$ fixes.
The function $p(x)$ can be found in terms of Jacobi elliptic functions,
XXX as shown in the Appendix,
but the expressions are particularly unhelpful.
The total successful mutational influx per generation is then given by $\int \rho \mu p(x) dx$,
and depends in a complicated way on the relationship of patch width and selection coefficients,
but still scales linearly with the population mutational influx $\rho \mu$.
Therefore, we define $p_f = \int p(x) dx$, and in a pinch will approximate this by $2 s_b A$,
which will be a good approximation if the width $\sqrt{A}$ is large relative to $\sigma/\sqrt{2s_m}$.

XXX We are interested in the case where a single mutation only takes over? XXX

@@

Now suppose that there are two patches, each of area $A$, of good environment separated by distance $R$,
and that the allele has begun to fix in the first, but has not yet appeared in the second.
As above, any copy of the new allele that appears at location $x$ relative to the second, unadapted patch 
has probability approximately $p(x)$ of fixing locally in this patch,
whether the allele is a new mutation or is descended from the occupants of the first patch.
Due to mutation--selection balance there will be a small number of alleles descended from those of the first patch (``migrants'') 
present in the regions where they are disadvantageous;
the equilibrium frequency may again be expressed in terms of Jacobi elliptic functions (Appendix XXX, citeXXX),
although the equations are again not particularly helpful.
However, we do know that if $q(r)$ is the expected proportion present at mutation--selection balance at distance $r$ from the first patch
in the absence of the second patch, that for large $r$ the proportion decays exponentially, $q(r) \approx C \exp( -|\sqrt{s_m} r / \sigma|)$,
where $C$ is a constant depending on the structure of the populations and the selection coefficients.

@@

>>>>

\section{Methods}
\subsection{Introduction to a patchily constructed model}
\label{ss:patchyspace}

Consider a population spread uniformly across a landscape. Within this environment there are patches of
habitat that individuals are initially poorly adapted to, surrounded by large areas that our population is well adapted to.

It takes only a single mutational change to create an allele ($B$) that adapts an individual to the
poor habitat type. This mutational change occurs at a low rate of
$\mu$ per chromosome per generation. We assume that this initially rare new
allele has fitness $1+s_b$ relative to the unmutated type ($b$) in the ``poor'' habitat patches,
and assume that it has fitness $1+s_m$ when rare in the intervening areas, with
$s_m<0$. To simplify matters we assume that the disadvantage $s_m$ ensures that on the relevant timescale,
the allele is very unlikely to fix locally in the regions where it is at a disadvantage.

We are interested in the establishment of mutations in the ``poor'' patches by either
migration or mutation, and so we are only interested in whether the allele
can escape initial loss by drift when rare. Thus we do not have to
specify the the fitness of the homozygote, only that its
fitness is such the dynamics of the  allele when it is rare are
determined by the fitness of the heterozygote. This allows us to treat
a diploid model as being essentially haploid for our purposes. \gc{TRUE?}

For the sake of simplicity we will assume a constant haploid population density $\rho$, 
across both types of habitat (we will return to discuss this equal
density point later). We further assume that the
variance in offspring number is $1$, and that the mean squared distance between parent and child is $\sigma^2$ (the dispersal variance).

As only a single mutation is required to adapt individuals to the
poor habitat patch, subsequent mutations that arise after an allele
becomes established at its equilibrium frequency on the patch gain no selective
benefit. Similarly an allele introduced into a patch by migration will
not be favored by selection to spread, if the patch has already been
colonized. Thus mutations selective exclude each other from
patches, over short-time scales, and they will only slowly filter
together over longer time scales by drift.

\paragraph{Previous work} We will make use of a number previous
results from the literature. \citet{slatkin1973geneflow}, working in
one dimensional deterministic framework, showed that if the
physical width of the patch is less than $2 \tan^{-1}
(\sqrt{|s_m/s_b|})$, then $B$ can not 
become stably established within the patch, as the local adaptation prevented by
migrational swamping \citep[see also][ for a review]{Lenormand}.
 \citet{barton1987establishment}, again working in one dimension,
 showed that this critical patch size also held in a stochastic
 framework \citep[see also the work of][]{Polk}. 
For patches above this size, \citet{barton1987establishment}
 founds that mutations appearing further than a few multiples of
 $\sigma/\sqrt{2|s_m|}$ away from such a patch have vanishingly small
 probability of becoming established within the patch. 
Barton also showed that mutations appearing within the patch have probability 
strictly less but approaching that $2s_b$ of establishment as the size
of the patch increased. Thus for large enough patches the probability
of establishment of a locally favorable allele approaches the standard
establishment probability ($2s_b$) for a panamitic population
\citep[][]{haldane, fisher}, a result that also (generally) holds for
a geographically spread population experiencing a uniform selection
pressure \citep{Maruyama, cherry}.

%Maruyama: On the fixation probability of mutant genes in a subdivided population*

%To make things tractable, we assume that the distances between the patches are large relative to migration distance $\sigma$,
%and that the disadvantage $s_m$ ensures that on the relevant timescale,
%the allele is very unlikely to fix locally in the regions where it is at a disadvantage.

%A similar situation of interest is the case when the species occupies a continuous range, but the selective pressure itself is patchy.
I%f the areas where the new allele is advantageous are relatively isolated,
%we might hope to apply the results above to such a situation.
%For this purpose, suppose that instead of a uniform habitat,
%the selective pressure is such that the new alleles are only advantageous in isolated locations, 
%and are disadvantageous in the intervening region.
%As before, we assume that the new allele has fitness $1+s_b$ relative to the unmutated type in the ``good'' patches,
%and assume that it has fitness $1+s_m$ in the intervening areas, with $s_m<0$.
%To make things tractable, we assume that the distances between the patches are large relative to migration distance $\sigma$,
%and that the disadvantage $s_m$ ensures that on the relevant timescale,
%the allele is very unlikely to fix locally in the regions where it is at a disadvantage.
%With these assumptions as justification, we will move to a model of discrete demes exchanging migrants,
%where ``migration'' occurs through the rare, transient passage of the new allele through regions where it is disadvantageous.

%equal deme sizes?  general migration rates?

\subsubsection{Mutational influx}
\label{ss:patchymutation}
Consider first a single, isolated poor habitat patch of area $A$ in
which no $B$ allele has yet become established. As
we are interested in local adaptation that can occur we will assume
that our patch is considerably larger than the cutoff for local
establishment, i.e. $\sqrt{A} \gg 2 \tan^{-1} (\sqrt{|s_m/s_b|})$.

Let $p(x)$ be the probability that a new allele arising at location
$x$ away from the patch fixes within the poor habitat patch.
The function $p(x)$ can be found in terms of Jacobi elliptic functions,
XXX as shown in the Appendix, but the expressions are particularly unhelpful.
The total successful mutational influx per generation is then given by $\int \rho \mu p(x) dx$,
and depends in a complicated way on the relationship of patch width and selection coefficients,
but still scales linearly with the population mutational influx $\rho \mu$.
Therefore, we define $p_f = \int p(x) dx$.  If the width of the patch, $\sqrt{A}$, is large
relative to $\sigma/\sqrt{2s_m}$ then a reasonable approximation is to
set $p_f \approx 2 s_b A$. This follows from the fact that if the
patch is large than the majority of the mutations that could become
established within the patch, arose within the patch.

The rate at which mutations arise and colonize the patch is on the order of $\rho \mu
p_f$ generations, which will approximately be $2 s_b \rho \mu$. If
this rate is low,  then the waiting time till a mutation arises that
will become locally established within the patch is exponentially
distributed with mean $2 s_b \rho \mu$.  Assuming that once a mutation
becomes established
 it quickly reaches its equilibrium frequency across the patch, the
 expected time it takes for our patch to become colonized by the $B$
 allele is $1/(2 s_b \rho \mu)$.

@@

Now suppose that there are two patches, each of area $A$, of the poor
habitat type separated by distance $R$. Consider the case where the
allele has arisen and become established in the first patch, but has not yet appeared in the second.
As above, any copy of the new allele that appears at location $x$ relative to the second, unadapted patch 
has probability approximately $p(x)$ of fixing locally in this patch,
whether the allele is a new mutation or is descended from the occupants of the first patch.
Due to mutation--selection balance there will be a small number of alleles descended from those of the first patch (``migrants'') 
present in the regions where they are disadvantageous;
the equilibrium frequency may again be expressed in terms of Jacobi elliptic functions (Appendix XXX, citeXXX),
although the equations are again not particularly helpful.
However, we do know that if $q(r)$ is the expected proportion present at mutation--selection balance at distance $r$ from the first patch
in the absence of the second patch, that for large $r$ the proportion decays exponentially, $q(r) \approx C \exp( -|\sqrt{s_m} r / \sigma|)$,
where $C$ is a constant depending on the structure of the populations and the selection coefficients.

@@

We will assume that once a mutation escapes loss and becomes
established in the poor habitat patch it reaches its equilbrium frequency very rapidly.
